Practical Optional Types for Clojure

### 2015-02-23 formal-model.tex
compact MM rules
let-aliasing rules
polish do/isa in section 3

Syntax is given in figure~\ref{main:figure:termsyntax}. Expressions include variables, values,
application, abstractions, conditionals, do (sequencing) expressions and let expressions.
We model throwable errors without a catching mechanism. All binding forms introduce fresh
variables.

Values include booleans, \nil{}, class literals, keywords, constants and closures.
\false{} and \nil{} are the only logically false values.

@@

Typing rules are in figure~\ref{main:figure:typingrules}.
T-Error neither returns logically true or false as it does not return normally.
T-Do uses the propositions learnt in the first expression to type check the second.
Other than compensating for the extra logically false value \nil{}, the rest are standard.

Figure~\ref{main:figure:standardopsem} contains the operational semantics, which are standard
except for \nil{} being false.

@@

\begin{figure*}
  \footnotesize
$$
\begin{altgrammar}
  \expd{}, \e{} &::=& \x{}
                      \alt \v{} 
                      \alt {\comb {\e{}} {\e{}}} 
                      \alt {\abs {\x{}} {\t{}} {\e{}}}
                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
                      \alt {\trdiff{\doexp {\e{}} {\e{}}}}
                      \alt {\letexp {\x{}} {\e{}} {\e{}}}
                      %\alt {\errorvalv{}}
                &\mbox{Expressions} \\
  \v{} &::=&          \singletonmeta{}
                      \alt {\const{}}
                      \alt {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
                &\mbox{Values} \\
  {\const{}}           &::=& \classconst \alt \throwconst

                &\mbox{Constants}\\
  \s{}, \t{}    &::=& \Top 
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt
                      {\ArrowOne {\x{}} {\t{}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}}
                      \alt {\Value \singletonmeta{}} 
                      \alt \trdiff{\class{}}
                &\mbox{Types} \\
  \singletonallsyntax{}
                \\ \\
  \occurrencetypingsyntax{}\\
  \openvsyntax{}
  %\\
  %\classliteralallsyntax{}
\end{altgrammar}
$$
\caption{Syntax of Terms, Types, Propositions and Objects}
\label{main:figure:termsyntax}
\end{figure*}

@@

    {\BDo}

@@

\subsection{\doliteral{}}

Assertions in Clojure often expand to simple exceptional control flow reasoning.

\begin{minted}{clojure}
(fn [x :- (U nil Number)] {:pre (number? x)} (inc x))
\end{minted}

@@

\begin{minted}{clojure}
(fn [x :- (U nil Number)]
  (do (if x (throw (new Exception)) nil)
      (inc x)))
\end{minted}

With an elegant extension in T-Do (figure~\ref{main:figure:typingrules}) we can follow this reasoning
to infer \mintinline{clojure}{(inc x)} is safe. The main insight is we can assume the
disjunction of the propositions of the first \doliteral{} argument while checking the second.

To sketch how this example is checked, first T-Abs type checks the body under the 
environment \propenv{} = {\isprop{\Union{\Nil}{\Number{}}}{\x{}}}.
T-Do then extracts the propositions of its first argument:
T-If tells us \thenprop{\prop{1}} = {\orprop{\botprop{}}{\botprop{}}} and 
\elseprop{\prop{1}} = {\orprop{\botprop{}}{\notprop{\falsy{}}{\x{}}}}
using T-Error, T-Nil and T-Subsume.
Now {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}} = {\notprop{\falsy}{\x{}}}
and we check \mintinline{clojure}{(inc x)} in the extended environment 
\ma{{\propenv{},{\notprop{\falsy}{\x{}}}}}
and by L-Update we know {\isprop{\Number}{\x{}}}
so T-Local says \mintinline{clojure}{x} is a \Number.

\subsection{\isaliteral{}}

One interesting Clojure function we modeled is \isaliteral.
It takes two arguments, and returns \true{} if they are the same (equality mode)
or if they are both classes and the left is a subtype of the right (subclassing mode),
otherwise returns \false{}, as described in figure~\ref{main:figure:isasemantics}.

@@

\begin{figure}
  \footnotesize
  \begin{mathpar}
    \isaopsemfigure{}

    \BIsA{}
  \end{mathpar}
  \caption{\isaliteral{} semantics}
  \label{main:figure:isasemantics}
\end{figure}

@@

\begin{figure}
  \footnotesize
  \getmethodfigure{}
  \begin{mathpar}
    \BDefMulti{}
    \BBetaMulti{}

    \BDefMethod{}
  \end{mathpar}
  \caption{Multimethod semantics}
  \label{main:figure:mmsemantics}
\end{figure}

@@

\begin{figure}
  \footnotesize
  \begin{mathpar}
    {\TNewStatic}
    {\TFieldStatic}

    {\TMethodStatic}
  \end{mathpar}
  \caption{Java Interoperability Typing Rules}
  \label{main:figure:javatyping}
\end{figure}

@@

\input{constant}
\input{update}

{\convertjavatypefigure{figure}}

>>>>

Syntax is given in figure~\ref{main:figure:termsyntax}. Expressions include variables, values,
application, abstractions, conditionals, do (sequencing) expressions and let expressions.
All binding forms introduce fresh variables.

Values include booleans, \nil{}, class literals, keywords (written with a : prefix like \mintinline{clojure}{:a}), 
constants and closures.  \false{} and \nil{} are logically false; all other values are logically true.

@@

Figures ~\ref{main:figure:typingrules} and~\ref{main:figure:standardopsem}
contain the typing rules and operational semantics.
They compensate for the extra logically false value \nil{}, otherwise they are standard.

@@

\begin{figure*}
  \footnotesize
$$
\begin{altgrammar}
  \expd{}, \e{} &::=& \x{}
                      \alt \v{} 
                      \alt {\comb {\e{}} {\e{}}} 
                      \alt {\abs {\x{}} {\t{}} {\e{}}}
                      \alt {\ifexp {\e{}} {\e{}} {\e{}}}
                      %\alt {\trdiff{\doexp {\e{}} {\e{}}}}
                      \alt {\letexp {\x{}} {\e{}} {\e{}}}
                      %\alt {\errorvalv{}}
                &\mbox{Expressions} \\
  \v{} &::=&          \singletonmeta{}
                      \alt {\const{}}
                      \alt {\closure {\openv{}} {\abs {\x{}} {\t{}} {\e{}}}}
                &\mbox{Values} \\
                \constantssyntax{}\\
  \s{}, \t{}    &::=& \Top 
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt
                      {\ArrowOne {\x{}} {\t{}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}}
                      \alt {\Value \singletonmeta{}} 
                      \alt \trdiff{\class{}}
                &\mbox{Types} \\
  \singletonallsyntax{}
                \\ \\
  \occurrencetypingsyntax{}\\
  \openvsyntax{}
  %\\
  %\classliteralallsyntax{}
\end{altgrammar}
$$
\caption{Syntax of Terms, Types, Propositions and Objects}
\label{main:figure:termsyntax}
\end{figure*}

@@

    %{\BDo}

@@

\subsection{Reasoning about Exceptional Control Flow}

Clojure assertions expand to simple exceptional control flow.

\begin{minted}{clojure}
(fn [x :- (U nil Number)] 
  (assert (number? x))
  (inc x))
\end{minted}

@@

\begin{minted}{clojure}
(fn [x :- (U nil Number)]
  (do (if (number? x) nil (throw (new Exception)))
      (inc x)))
\end{minted}

We extend our model with sequencing expressions and errors, where {\errorvalv{}}
is equivalent to a Clojure throw expression.

\smallskip
$
\begin{altgrammar}
  \e{} &::=& \ldots \alt {\errorvalv{}} \alt {\doexp {\e{}} {\e{}}} &\mbox{Expressions} 
\end{altgrammar}
$

\smallskip

B-Do simply evaluates its arguments sequentially and returns the right argument.
Since errors are not values, we define error propagation semantics
like BE-Do1 (figure~\ref{appendix:figure:errorstuck} for the full rules).

\begin{mathpar}
    {\BDo}

\infer [BE-Error]
{}
{ \opsem {\openv{}} 
         {\errorvalv{}}
         {\errorvalv{}}}

\infer [BE-Do1]
{ \opsem {\openv{}} {\e{1}} {\errorvalv{}} }
{ \opsem {\openv{}} {\doexp{\e{1}}{\e{}}} {\errorvalv{}}}
\end{mathpar}

Our main insight is as follows: 
if the left argument reduces to a value, then it is either logically true or false.
Occurrence typing gives exactly the propositions for both cases, so their disjunction
can be assumed for the right argument.
T-Do formalises this intuition.

\begin{mathpar}
    {\TDo}

    {\TError}
\end{mathpar}

The introduction of errors, specifically as an expression that is neither true nor false, 
makes our insight interesting.
For example, in checking \mintinline{clojure}{(do (throw (Exception.)) (inc :a))}
by T-Error we assume \orprop{\botprop{}}{\botprop{}} in checking \mintinline{clojure}{(inc :a)},
which is a contradiction by the logic system (L-False, figure~\ref{appendix:figure:proofsystem})
which allows us to assume \mintinline{clojure}{(inc :a)} is unreachable (and it clearly is by BE-Do1).

This works similarly with our motivating example. Intuitively, when evaluating
\mintinline{clojure}{(inc x)} we know \mintinline{clojure}{(assert (number? x))} has evaluated to a value,
so we assume \mintinline{clojure}{x} is a number.
In detail, the example is checked first by T-Abs, which checks the body with the 
environment \propenv{} = {\isprop{\Union{\Nil}{\Number{}}}{\x{}}}.
Since the body is a \doliteral{},
T-Do extracts the propositions of the left argument, an if:
T-If says 
\thenprop{\prop{1}} = {\orprop{\botprop{}}{\notprop{\falsy{}}{\x{}}}}
by T-Nil and T-Subsume and
\elseprop{\prop{1}} = {\orprop{\botprop{}}{\botprop{}}} by T-Error.
Now {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}} = {\notprop{\falsy}{\x{}}}
and we check \mintinline{clojure}{(inc x)} in the extended environment 
\ma{{\propenv{},{\notprop{\falsy}{\x{}}}}}.
By L-Update (figure~\ref{appendix:figure:proofsystem}) we know {\isprop{\Number}{\x{}}},
so T-Local checks \mintinline{clojure}{x} as a \Number.

\subsection{Multimethod dispatch mechanism: \isaliteral}

Before tackling multimethods we model their core dispatch mechanism
\isaliteral{}, a subclassing test for classes, otherwise an equality test.

\begin{minted}{clojure}
(isa? Keyword Object) ;=> true
(isa? Object Keyword) ;=> false
(isa? :a :a) ;=> true
(isa? :a :b) ;=> false
(isa? :a Keyword) ;=> false
(isa? (class :a) Keyword) ;=> true
\end{minted}

B-IsA with the metafunction \isaopsemliteral{} give the operational semantics.

\smallskip
$
\begin{array}{ll}
  \vcenter{\hbox{\BIsA{}}}
  &
  \vcenter{\hbox{\isaopsemfigure{}}}
\end{array}
$
\smallskip

For the typing rules, we add rules for type checking keyword and class literals
as singleton types, and define T-IsA, which infers the propositions of an \isaliteral{} 
expression by passing
 the \emph{object} of the left argument and the \emph{type}
of the right argument to \isacompareliteral{}.

\smallskip
$
\begin{array}{ll}
\vcenter{\hbox{
    \TIsA{}}} &

  \vcenter{\hbox{$
    \begin{array}{l}
      {\TKw}\\\\
      {\TClass}
    \end{array}$}}
\end{array}
$
\smallskip

\begin{figure*}
  \begin{mathpar}
    \isapropsfigure{}
  \end{mathpar}
  \caption{\isaopsemliteral Proposition set Metafunction}
\end{figure*}

\begin{mathpar}
    {\TKw}

    {\TClass}

    \isapropsfigure{}

    \TIsA{}
\end{mathpar}

@@

@@

\begin{figure}
  \footnotesize
  \getmethodfigure{}
  \begin{mathpar}
    \BDefMethod{}
    \BDefMulti{}
    \BBetaMulti{}
  \end{mathpar}
  \caption{Multimethod and method resolution semantics}
  \label{main:figure:mmsemantics}
\end{figure}

@@

\begin{figure*}
  \footnotesize
  \begin{mathpar}
    {\TNewStatic}

    {\TFieldStatic}
    {\TMethodStatic}
  \end{mathpar}
  \caption{Java Interoperability Typing Rules}
  \label{main:figure:javatyping}
\end{figure*}

@@

{\footnotesize
\input{constant}
\input{update}
}

{\convertjavatypefigure{figure}{\footnotesize}}
